ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-08 notebooks/circles.qmd
code for tables mostly sorted

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="Moons",
    n_individuals=5,
    target=0, factual=1,
    initialization=:identity,
    converge_when=:generator_conditions,
)
CSV.write(joinpath(output_path, "circles_benchmark.csv"), bmk())
```

>>>>

bmk = benchmark(
    counterfactual_data; 
    models=model_dict, 
    generators=generator_dict, 
    measure=measures,
    suppress_training=true, dataname="Circles",
    n_individuals=5,
    target=0, factual=1,
    initialization=:identity,
    converge_when=:generator_conditions,
)
CSV.write(joinpath(output_path, "circles_benchmark.csv"), bmk())
```

### 2023-05-08 notebooks/mnist.qmd
code for tables mostly sorted

```{julia}
Random.seed!(1234)

@@

# Training params:
T = 100
opt = Flux.Optimise.Adam(0.01)
```

```{julia}
# Search:
generic_generator = GenericGenerator(opt=opt)
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=1.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-1]      # penalty strengths
```

@@

```{julia}
# Plot generated samples:
n_regen = 200
if _regen 
    for (mod_name, mod) in model_dict
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

>>>>

```{julia}
Random.seed!(2023)

@@

# Training params:
T = 100
opt = Flux.Optimise.Adam(0.1)
```

```{julia}
# Search:
generic_generator = GenericGenerator(opt=Flux.Optimise.Descent(1.0))
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=2.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

@@

```{julia}
# JEM parameters:
𝒟x = Uniform(-1.0,1.0)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,1e-1]      # penalty strengths
```

@@

```{julia}
# Plot generated samples:
n_regen = 500
if _regen 
    for (mod_name, mod) in model_dict
        K = length(counterfactual_data.y_levels)
        input_size = size(selectdim(counterfactual_data.X, ndims(counterfactual_data.X), 1))
        𝒟x = Uniform(extrema(counterfactual_data.X)...)
        𝒟y = Categorical(ones(K) ./ K)
        sampler = ConditionalSampler(𝒟x, 𝒟y; input_size=input_size, prob_buffer=0.0)
        opt = ImproperSGLD()
        f(x) = logits(mod, x)

### 2023-05-08 notebooks/tables.Rmd
code for tables mostly sorted

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures,
    .(value=sprintf("%1.2f\n(%1.2f)", mean(value), sd(value))),
    .(dataname, generator, model, variable, source)
]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
tab <- dcast(tab, model + variable + generator ~ dataname)
kbl(tab, align = "c") %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, valign = "top")
```

>>>>

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="real",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,min_val:=NULL]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for real-world datasets. Standard deviation in parentheses. \\label{tab:results-real}", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_real.tex")
```

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="synthetic",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,min_val:=NULL]
tab[generator == "eccco", generator:="ECCCo"]
tab[generator == "revise", generator:="REVISE"]
tab[generator == "generic", generator:="Wachter"]
tab[generator == "wachter", generator:="Wachter"]
tab[generator == "greedy", generator:="Schut"]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for synthetic datasets. Standard deviation in parentheses. \\label{tab:results-synth}", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  # kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "Moons" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_synth.tex")
```

### 2023-05-08 paper/paper.tex
work on table
code for tables mostly sorted

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\title{ECCCos from the Black Box: Faithful Counterfactual Explanations through Energy Based Modeling and Conformal Prediction}

@@

\begin{abstract}
  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models but they are not unique. To address the need for plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate. Consequently, the generated explanations may look plausible to humans but not necessarily faithfully describe the behaviour of the black-box model. We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits. To do so, \textbf{ECCCo} leverages recent advances with respect to energy-based models and conformal inference. Through extensive empirical studies, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness. Our findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs yield more plausible explanations than pure discriminative models.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_benchmark.png}
  \caption{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:mnist-benchmark}
\end{figure}

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{import}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}

@@

\title{ECCCos from the Black Box:\\
Faithful Explanations through\\
Energy-Constrained Conformal Counterfactuals}

@@

\begin{abstract}
  Counterfactual Explanations offer an intuitive and straightforward way to explain black-box models but they are not unique. To identify the most plausible explanations, existing work has primarily relied on surrogate models to learn how the input data is distributed. This effectively reallocates the task of learning realistic representations of the data from the model itself to the surrogate. Consequently, the generated explanations may look plausible to humans but not necessarily faithfully describe the behaviour of the black-box model. We formalise this notion of faithfulness through the introduction of a tailored evaluation metric and propose a novel algorithmic framework for generating \textbf{E}nergy-\textbf{C}onstrained \textbf{C}onformal \textbf{Co}unterfactuals that are only as plausible as the model permits. Through extensive empirical studies involving multiple synthetic and real-world datasets, we demonstrate that ECCCos reconcile the need for plausibility and faithfulness. In particular, we show that it is possible to achieve state-of-the-art plausibility for any black-box model with gradient access without the need for surrogate models. To do so, \textbf{ECCCo} relies solely on properties defining the black-box model itself by leveraging recent advances in energy-based modelling and conformal inference. The empirical findings also highlight that black-box models that are trained jointly to discriminate outputs and generate inputs tend to yield more plausible explanations than pure discriminative models. Our framework is intuitive, flexible and open-sourced. By highlighting the need for faithfulness in the context of Counterfactual Explanations, we believe that in the short term, our work will enable researchers and practitioners to better distinguish trustworthy from unreliable models. We anticipate that ECCCo can serve as a baseline for future research directed at providing plausible but faithful Counterfactual Explanations. 
\end{abstract}

@@

See Table~\ref{tab:results-real}

\import{contents/}{table_synth.tex}

\import{contents/}{table_real.tex}
